Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2020-02-15 ms/second-review-responses.tex

\vskip 4em

\pagebreak
\setcounter{page}{1}

%%%%%%%%%%%%%%
\reviewersection{Decision Letter}

\begin{quote}
    It is most important that you address the following in a revised manuscript:
    \begin{enumerate}
        \item in response to reviewer 3, add text to help clarify how the reproduction scheme / dispersal model differs from models that do not induce the bimodality in the distance distribution the reviewer points out (see my comments below as well); 
        \item Respond to the concern about the definitions of sigma and neighborhood size; 
        \item Reviewer 2 has several minor comments that would be straightforward and helpful to address/add depth into your discussion (e.g., comments 2, 4, 5). 
        \item The other minor comments from each reviewer. 
    \end{enumerate}
\end{quote}

Thanks very much for the useful input.
The inaccuracy in describing dispersal under the model was our mistake,
and we have added more careful discussion of this point (see below).
We have modified the manuscript to incorporate all remaining feedback,
as enumerated below,
and think this has improved the manuscript.

\reviewersection{AE Comments}

\begin{quote}
The point of R3 on the bimodality in the realized dispersal distance in your model is interesting. You can make this more explicit in the paper and describe how there are realistic biological scenarios that motivate such a model (many/most plants) and demarcate how well you think you can generalize to others (the new appendix already helps in that regard). It strikes me your model has relevance for patrilocal/matrilocal systems where the paternal/maternal parent-offspring distances differ - though your model is hermaphroditic so it's an imperfect analogy. Perhaps, to explore more typical animal mating patterns, you can quickly check the impact for one or two outcomes of changing the simulation such that offspring are dispersed around the midpoint of the parental origins, as this would remove the bimodality. That said, please be sure we respect the work done is extensive and don't mean to invite/require a complete re-working. The core tension is between ensuring your results are generalizable if that is what you wish to claim; versus specific to a particular model that is biologically motivated. It's particularly key for the material on GWAS, where you are explicitly trying to address human GWAS results: those sections will be read and taken at face value by a large community even if this simulation scheme introduces some discrepancies from what might more realistically emerge. So at a minimum, some caveats directly in the GWAS discussion section would help.
\end{quote}

Thank you for your comments. We have responded to each reviewer below. In response to reviewer 2 we have added several caveats to the GWAS section noting other sources of error such as allele frequency and LD variation among populations. Reviewer 3 raised two good points on our definition of neighborhood size and the interpetation of $\sigma$ -- first, that $\sigma$ refers to one- rather than two-dimensional distances between parent and offspring (this is consistent with previous theoretical work, and we have corrected our definition in the text), and second, that the use of separate draws for mate selection and dispersal means the total distance traveled by a locus along a pedigree link is greater for ``paternal'' than ``maternal'' material (taking ``maternal'' to mean the parent from which dispersal is calculated). Because our simulated genome is autosomal and individuals hermaphroditic and non-selfing, a locus will experience a mix of ``maternal'' and ``paternal'' transmission events along its genealogy. So we see this as a scaling issue rather than one of a ``bimodal'' dispersal kernal per se -- the expected axial displacement of a locus along a pedigree link is $\sqrt{3/2} \sigma$, rather than $\sigma$.
(Note that although the parent-offspring displacement is a mixture of mean-zero Gaussians
with respect to a bimodal distribution of standard deviations,
the norm of any such mixtures is never itself bimodal.)
% However we note that in terms of summary statistics the effects of scaling $\sigma$ by $\sqrt{3/2}$ are weaker than those generated by factors like changing landscape resolution in lattice vs. continuous models (e.g. Appendix 1), so we don't think this significantly changes our interpretation of model outputs.
We have added a paragraph discussing these points (while trying to keep out of the weeds),
to \llname{ll:sigma_effective}.

@@

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
Only minor changes were needed, and they were made. This study neatly summarises many biases arising from unmodelled continuous space (pertinent for real studies), and I would not extend it by further analyses. When polishing the final version, the authors can make some of their messages stronger. For example, Figure 7 and the accompanying text may give the impression that we are doing pretty well with smc++. It is only when panel B is seen one can appreciate the degree of uncertainty. So, in a way, Figure S7 is more telling as to how much we can trust our results in real analysis. After all, in reality, as authors mention, we only have one sample set, not replicates.
\end{quote}

Thank you for your comments. While we agree that our analysis found problems with demographic inference under isolation by distance, since this manifests as higher variance in predictions rather than a straightforward bias as we expected (i.e. a reduction in recent inferred Ne), we have chosen to keep the somewhat more nuanced presentation of results in that section used in a previous version,
although we added an additional note about the between-replicate variance to the discussion.
\llname{ll:variance_note}

@@

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
The authors have done a commendable job of responding to the reviewers' comments. The extended discussion and additional supplemental figures helps with the clarity of the manuscript. The results for the comparison with the discrete stepping-stone model were useful - though see below. The authors have also provided sufficient detail justifying their choice of model parameters.
\end{quote}

\begin{quote}
I don't think I have any critical remaining issues or suggestions, but I did have a couple of thoughts. The authors might just want to think about them.
\end{quote}

Thanks for the encouraging words and helpful points,
as well as for your mindfulness of the burden of additional analyses.

@@

\begin{point}{}
I appreciate the appendix comparing summary statistics in the continuous and stepping-stone models, although I must confess I am now even more uncertain than I was before about which models are more appropriate. I get that the coalescent models make assumptions that are not directly interpretable in terms of observable parameters (although are ``interactions between geographically distant parts of the species range'' really biologically unrealistic?). This forward model is more interpretable, in that sense, but does it actually produce more realistic data? I guess that's a question for future work, but that question was one of the main things I came away from this paper with.
\end{point}

\reply{
This is a good point. We have added additional discussion noting that while our continuous model has some advantages over discrete models, the best model for any empirical system will inevitably depend on the life history of the organism in question. \llname{bestmodelcaveat}
}

@@

\begin{point}{\revref}
 Is Khera et al (2018) really the right reference for the idea of polygenic scores? They have been around for a lot longer than that.
\end{point}

\reply{Yes this was an oversight. We now cite Wray et al. 2007 for individual disease risk prediction from GWAS estimates and the 
first, to our knowledge, PGS application (International Schizophrenia Consortium, 2009).
}

\begin{point}{\revref}
 I don't know that the results of Berg and Sohail et al. are indicating a fundamental issue with PC correction, rather than a problem with meta-analysis.
\end{point}

\reply{
    We think \citet{Berg2018} shows that population structure leaves signals in GWAS that resemble those of polygenic selection, and given that this was most apparent in GIANT data employing PCA corrections, indicates a problem with the GWAS that make up the GIANT study at least as much as any issues with meta-analysis. 
}

@@

\begin{point}{\llname{ldcaveats}}
 Is the result of Martin et al. really due to population stratification? There seem to be a number of factors affecting transferability of PRS (LD differences, effect size heterogeneity, differences in causal variants etc..), but I don't know that anyone thinks residual structure is the major one, particularly for eg. UKB summary stats.
\end{point}

\reply{This is a good point -- differences in allele frequency and LD structure certainly have significant effects on PRS transferability (at least when the GWAS is based on SNP assays rather than whole-genome sequencing). We think the population structure issue is important to point to here because it will not be solved by simply generating more detailed genomic data from existing samples, but have added a caveat noting these other important potential sources of error at the head of that sentence. \llname{ldcaveats}
}

\begin{point}{``population genetics flavor'' \revref}
I don't understand what this means?
\end{point}

\reply{
We meant to contrast this sort of ``single-species'' model,
common in population genetics,
to a more ``ecological'' model involving explicit modeling or resources and/or other species.
We've reworded this. \revref
}

@@

\reply{
thanks for catching this, corrected. 
}

%%%%%%%%%%%%%%
\reviewersection{3}

\begin{quote}
I am fairly pleased overall with the revised version of this article. I commend the authors for the substantial amount of work conducted. This contribution now convincingly demonstrates how important it is to properly account for population structure when individuals are distributed along a spatial continuum.
\end{quote}

Thanks. We are also fairly pleased with it, overall.
\plr{this is probably too much =)}

@@

\begin{point}{}
A couple of minor points about the forward model are still puzzling me though. First of all, in one dimension, the variance of $L(o)$, the random variable corresponding to the offspring's spatial position, given $l(p)$, the parent's position, is equal to $E((L(o)-l(p))^2):=\sigma^2$. Hence, $\sigma$ is here the square root of the expected squared Euclidean distance between parents and offspring. In general, assuming isotropic migrations, $\sigma$ is the square root of the expected squared Euclidean distance taken along any dimension. It seems to me that this definition is distinct from the mean parent-offspring distance, as stated by the authors.
\end{point}

\reply{
Good point here.
We did intend to define $\sigma$ as is usual in the literature,
so have corrected the wording
and now direct readers to \citep{Rousset1997} for discussion of how these parameters are defined in one- and two-dimensional habitats. \llname{axialdist}
}

@@

\begin{point}{}
More importantly, I am still a bit confused about the reproduction scheme. My understanding is that, once mating takes place, the two parents involved (noted as P(1) and P(2)) each produce a Poisson number of offspring. The location of each offspring is determined by a Gaussian distribution of covariance matrix $\sigma^2 I$, centered on the location of the parent that ``produced' this offspring. Let $O(1,i)$ and $O(2,j)$ be the i-th and j-th offspring produced by $P(1)$ and $P(2)$ respectively. Now, if I am not mistaken, half of the genetic material of O(2,j) comes from P(1) (the other half coming from P(2)) while O(1,i) receives one chromosome from P(2) (and another one from its ``true' parent, i.e., P(1)). Hence, when it comes to the parent-offspring distance as measured from their genetic material, half of the chromosomes have a ``small' dispersal distance, equal to that of the parent-offspring distance, while the other half have a ``large' distance roughly equal to the parent-offspring distance plus the distance between mates.

It is important that the authors clarify this last point because most results presented in this study rely on Wright's neighborhood size, which derives from a model where dispersal of individuals and chromosomes follow well-defined spatial dynamics. A sensible comparison of neighborhood sizes requires that dispersal (along with effective population density) is quantified in a coherent way across different models.

Additionally, the neighborhood size as defined by the authors (see \llname{ll:nw_defn}) is a function of the census density rather than the effective density of individuals. I am therefore not fully convinced that the range of values for this parameter is comparable to that given in Table 1.
\end{point}

\reply{
    You are entirely correct here; apologies for the confusion.
    To clarify the definition of $\sigma$ and discuss these issues somewhat,
    we have added a new paragraph at \llname{ll:sigma_effective}.
    As we discuss there, we still calculate $N_W = 4 \sigma^2 \rho$,
    where $\sigma$ is the mean mother-child dispersal distance
    (not including the distance between mates)
    and $\rho$ is the census population density.
    This is not the quantity that is predicted
    to best describe geographic patterns of genetic variation,
    which would use \emph{effective} versions of both $\sigma$ and $\rho$.
    This does not bother us for a few reasons.
    First, we do not attempt to match statistics in our simulations
    to theoretical results (e.g., the Wright--Malec\'ot formula).
    Instead, we use $N_W$ as a proxy for ``strength of spatial structure'',
    and in these simulations, the ``effective'' $N_W$ certainly has a monotonic
    relationship with the $N_W$ we compute.
    The remaining question is whether, as the reviewer asks,
    the actual values of $N_W$ are comparable to those computed in the wild.
    This depends, of course, on how those values were estimated in wild populations.
    In our opinion, a straightforward estimate of $N_W$
    using observed dispersal distances and census density or reproductive adults
    (similar to how we compute $N_W$ for simulations)
    is likely more robust and reflective of modern demography
    than most methods to estimate ``effective'' vesions.

    Furthermore, our imprecisions roughly cancel out.
    As now noted in the text, an average of one and two dispersal distances
    is a better model for the mean parent-offspring distance;
    this would give us a value of $\sqrt{3/2} \sigma$,
    so by using only $\sigma$ we have lowered $N_W$ by a factor of about 2/3.
    However, effective population density is lower than census density
    by a factor of also around 2/3,
    which would cancel the first.
    So, we are fairly confident that our values of $N_W$ match the theoretical versions
    to within the precision we require (which is only order-of-magnitue, really).

These are an important point set of points on relating neighborhood sizes from different models which we have aimed to clarify. In our simulation the x axis displacement of a locus along any link in the pedigree is normally distributed with standard deviation $\sigma$ for the ``maternal'' material (i.e. deriving from the parent from which dispersal is calculated), but for the ``paternal'' parent it is the sum of two independent normal draws each with with standard deviation $\sigma$. Averaging over a random series of maternal and paternal transmissions (because our simulated genome is autosomal), variance in axial position across pedigree links is then $(3/2)\sigma^2$ and the expected displacement $\sqrt{3/2}\sigma$. So from the perspective of genes sampled at the present time our neighborhood size calculations are lower than those in models like those described in \cite{Rousset1997} by a factor of (3/2) because of differences in the breeding model.

### 2020-02-18 ms/spaceness.tex

\begin{abstract}
    Real geography is continuous,
    but standard models in population genetics are based on discrete, well-mixed populations. 
    As a result many methods of analyzing genetic data assume that samples are a random draw from a well-mixed population, but are applied to clustered samples from populations that are structured clinally over space.
    Here we use simulations of populations living in continuous geography 
    to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies. 
    We find that most common summary statistics have distributions that differ substantially from that seen in well-mixed populations, 
    especially when Wright's neighborhood size is less than 100 and sampling is spatially clustered. 
    Stepping-stone models reproduce some of these effects, but discretizing the landscape introduces artifacts which in some cases are exacerbated at higher resolutions.
        The combination of low dispersal and clustered sampling causes demographic inference from the site frequency spectrum to infer more turbulent demographic histories, but averaged results across multiple simulations were surprisingly robust to isolation by distance. 
    We also show that the combination of spatially autocorrelated environments and limited dispersal
    causes genome-wide association studies to identify spurious signals of genetic association 
    with purely environmentally determined phenotypes,
    and that this bias is only partially corrected by regressing out principal components of ancestry. 
    Last, we discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
\end{abstract}

@@

To study this, we have implemented an individual-based model in continuous geography 
that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation such as humans and \textit{Drosophila}, and output the full genealogy and recombination history of all final-generation individuals. 
We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
typically analyzed assuming discrete population models. 
We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
Last, we examine the impacts of continuous geography on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.

@@

% The best-studied approaches to population genetics in continuous space were developed by \citet{Wright1942,Wright1943} and \citet{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
depends on the chance that two geographically nearby individuals are close relatives
-- in modern terms, by the tension between
migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation by \citet{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
defined to be $N_W = 4 \pi \rho \sigma^2$, where $\sigma$ is the mean parent--offspring distance along each of the x and y axes and $\rho$ is population density (see \citep{Rousset1997} for further discussion of parameter defiitions in one- and two-dimensional habitats) \llabel{axialdist}. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species
-- even in human populations,
estimates range from 40 to over 5,000 depending on the population and method of estimation (Table~\ref{table:NStable}).
%\citet{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

@@

We simulated populations using the program SLiM v3.1 \citep{Haller2019}. 
Each time step consists of three stages: reproduction, dispersal, and mortality. 
To reduce the parameter space we use the same parameter, denoted $\sigma$, 
to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
As in previous work \citep{Rousset1997}, $\sigma$ is equal to the mean parent-offspring distance along each of the x and y axes.

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$.
Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates. 
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $d$ proportional to $\exp(-d^2 / 2 \sigma^2)$.

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$,
where $g$ is the Gaussian density with mean zero and standard deviation $\sigma$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, \llabel{ll:K_defn}
the probability of survival until the next time step for individual $i$ is
\begin{equation} \label{eqn:p_survival}
    p_i 
    =
    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is close to $K$, \llabel{ll:K_again}
and the mean lifetime is around $L$;
for more description see the Appendix. % \ref{apx:demographic_model}.

@@

Our demographic model includes parameters that control population density ($K$), 
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantitites to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations,
and used mean population density ($\rho$, census size divided by total area)
to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$.  \llabel{ll:nw_defn}
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the lifetime total number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period,
% subset the resulting table to include only the last timestep recorded for each individual,
and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citep{pyslim}.

@@

The effects of sampling on summary statistic estimates were summarized by testing for differences in mean (ANOVA, \citep{Rcore2018}) and variance (Levene's test, \citep{Fox2011}) across sampling strategies for each summary statistic.

@@

Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation. As currently implemented running simulations at neighborhood sizes more than 1,000 to coalescence is likely impractical, though running these models for more limited timescales and then ``recapitating'' the simulation using reverse-time simulation from the resulting tree sequence in msprime is possible \citep{haller2019treesequence}.  \llabel{runtimesxn}

@@

Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015,leblois2006genetics}. Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy systematically inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by around a factor of 2. We were surprised to see that both stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e., neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations. Inference in populations with neighborhood sizes over 20 was relatively unbiased unless samples were concentrated in the middle of the range (Figure \ref{fig:demography}). 
Although median demography estimates across many independent simulations were fairly accurate,
empirical work has only a single estimate to work with,
and individual model fits (Figure \ref{fig:demography_supplement})
suggest that spuriously inferred population size changes and bottlenecks are common, especially at small neighborhood sizes. 
As we will discuss below, most empirical estimates of neighborhood size, including all estimates for human populations, are large enough that population size trajectories inferred by these approaches should not be strongly affected by spatial biases created by dispersal in continuous landscapes. In contrast, \citet{Mazet2015} found that varying migration rates through time could create strong biases in inferred population trajectories from an $n$-island model with parameters relevant for human history, suggesting that changes in migration rates through time are more likely to drive variation in inferred $N_{e}$ than isolation by distance.

@@

We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
(note this is likely not the case for some modern human populations): 
run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations.
If significant associations with sample locality are observed after correcting for population structure, 
the method is sensitive to false positives induced by spatial structure. This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes). This approach has recently been taken with polygenic scores for UK Biobank samples in \citet{haworth2019apparent}, finding that scores are correlated with birth location even in this relatively homogenous sample. 
Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.

@@

One facet of spatial variation that we did not address in this study is the confounding of dispersal and population density implicit in the definition of Wright's neighborhood size. Our simulations were run under constant densities, but \citet{guindon2016demographic} and \citet{Ringbauer2017} 
have shown that these parameters are identifiable under some continuous models. Similarly, though the scaling effects of dispersal we show in Figure \ref{fig:sumstats} should occur in populations of any total size, other aspects such as the number of segregating sites are also likely affected by the total landscape size (and so total census size). \llabel{landscapesizecaveat} 
Indeed, our finding that stepping-stone models reproduce only some of the effects on genetic variation seen in a continuous-space model (Figure \ref{fig:grid_stats}) suggests that qualitatively similar models can produce different results dependent on the specific parameterizations used. While we believe our continuous model is a more appropriate depiction of evolution in a continuous landscape than lattice models, it is likely that some populations and breeding systems do more closely resemble a series of interconnected random-mating populations. As with all population models, the best approximation for any empirical system will depend on the natural history of the species in question. \llabel{bestmodelcaveat}
Much additional work remains to be done to better understand how life history and range size interact to shape genetic variation in continuous space, which we leave to future studies.

>>>>

\begin{abstract}
    Real geography is continuous,
    but standard models in population genetics are based on discrete, well-mixed populations. 
    As a result many methods of analyzing genetic data assume that samples are a random draw from a well-mixed population, but are applied to clustered samples from populations that are structured clinally over space.
    Here we use simulations of populations living in continuous geography 
    to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies. 
    We find that most common summary statistics have distributions that differ substantially from that seen in well-mixed populations, 
    especially when Wright's neighborhood size is less than 100 and sampling is spatially clustered. 
    Stepping-stone models reproduce some of these effects, but discretizing the landscape introduces artifacts which in some cases are exacerbated at higher resolutions.
    The combination of low dispersal and clustered sampling causes demographic inference from the site frequency spectrum to infer more turbulent demographic histories, but averaged results across multiple simulations
    revealed surprisingly little systematic bias.
    We also show that the combination of spatially autocorrelated environments and limited dispersal
    causes genome-wide association studies to identify spurious signals of genetic association 
    with purely environmentally determined phenotypes,
    and that this bias is only partially corrected by regressing out principal components of ancestry. 
    Last, we discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
\end{abstract}

@@

To study this, we have implemented an individual-based model in continuous geography 
that incorporates overlapping generations, local dispersal of offspring, and density-dependent survival. 
We simulate chromosome-scale genomic data in tens of thousands of individuals from parameter regimes relevant to common subjects of population genetic investigation, and output the full genealogy and recombination history of all final-generation individuals. 
We use these simulations to test how sampling strategy interacts with geographic population structure to cause systematic variation in population genetic summary statistics 
typically analyzed assuming discrete population models. 
We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. 
Last, we examine the impacts of continuous geography on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.

@@

% The best-studied approaches to population genetics in continuous space were developed by \citet{Wright1942,Wright1943} and \citet{Malecot1948}, who derived expressions for genetic differentiation in continuous space assuming Poisson distributed numbers of offspring and independent dispersal among individuals. 
The degree to which genetic relationships are geographically correlated
depends on the chance that two geographically nearby individuals are close relatives
-- in modern terms, by the tension between
migration (the chance that one is descended from a distant location)
and coalescence (the chance that they share a parent).
A key early observation by \citet{Wright1946} is that this balance is often nicely summarized by the ``neighborhood size'',
defined in two dimensions to be $N_W = 4 \pi \rho \sigma^2$,
where $\sigma^2$ is one half of the mean squared parent--offspring distance
and $\rho$ is population density
(see \citet{Rousset1997} for further discussion of parameter definitions in one- and two-dimensional habitats) \llabel{axialdist}. 
This can be thought of as proportional to the average number of potential mates for an individual (those within distance $2 \sigma)$,
or the number of potential parents of a randomly chosen individual. Empirical estimates of neighborhood size vary hugely across species
-- even in human populations,
estimates range from 40 to over 5,000 depending on the population and method of estimation (Table~\ref{table:NStable}).
%\citet{Maruyama1972} found that the rate of decline in genetic diversity in a two-dimensional continuous population approaches the random mating expectation when $\rho\sigma^2 > 1$, and proposed that this had the important implication that most population genetic expectations for randomly mating populations could be applied to continuously distributed populations with relatively little error.

@@

We simulated populations using the program SLiM v3.1 \citep{Haller2019}. 
Each time step consists of three stages: reproduction, dispersal, and mortality. 
To reduce the number of parameters we use the same parameter, denoted $\sigma$, 
to modulate the spatial scale of interactions at all three stages by adjusting the standard deviation of the corresponding Gaussian functions. 
Informally, we think of $\sigma$ as the ``dispersal distance'',
although only one of those stages is dispersal.

At the beginning of the simulation individuals are distributed uniformly at random on a continuous, square landscape. 
Individuals are hermaphroditic, and each time step, each produces a Poisson number of offspring with mean $1/L$.
Offspring disperse a Gaussian-distributed distance away from the parent with mean zero and standard deviation $\sigma$
in both the $x$ and $y$ coordinates. 
Each offspring is produced with a mate selected randomly from those within distance $3 \sigma$,
with probability of choosing a neighbor at distance $d$ proportional to the Gaussian density
with mean zero and standard deviation $\sigma$, which is $g(d) = \exp(-d^2 / 2 \sigma^2)/ (2 \sigma^2)$.

To maintain a stable population, mortality increases with local population density.
To do this we say that individuals at distance $d$ have a competitive interaction with strength $g(d)$.
Then, the sum of all competitive interactions with individual $i$ is $n_i = \sum_j g(d_{ij})$,
where $d_{ij}$ is the distance between individuals $i$ and $j$ and the sum is over all neighbors within distance $3 \sigma$.
Since $g$ is a probability density, $n_i$ is an estimate of the number of nearby individuals per unit area.
Then, given a per-unit carrying capacity $K$, \llabel{ll:K_defn}
the probability of survival until the next time step for individual $i$ is
\begin{equation} \label{eqn:p_survival}
    p_i 
    =
    \min\left( 0.95, \frac{1}{1 + n_i / (K (1+L))} \right) .
\end{equation}
We chose this functional form so that the equilibrium population density per unit area is close to $K$, \llabel{ll:K_again}
and the mean lifetime is around $L$;
for more description see the Appendix. % \ref{apx:demographic_model}.

@@

Our demographic model includes parameters that control population density ($K$), 
mean life span ($L$), and dispersal distance ($\sigma$).
However, nonlinearity of local demographic stochasticity causes
actual realized averages of these demographic quantities to deviate from the specified values
in a way that depends on the neighborhood size.
Therefore, to properly compare to theoretical expectations,
we empirically calculated these demographic quantities in simulations. 
We recorded the census population size in all simulations,
and used mean population density ($\rho$, census size divided by total area)
to compute neighborhood size as $N_W = 4 \pi \rho \sigma^2$.  \llabel{ll:nw_defn}
To estimate generation times, we stored ages of the parents of every new individual born across 200 timesteps, 
after a 100 generation burn-in, and took the mean. 
To estimate variance in offspring number, 
we tracked the lifetime total number of offspring for all individuals for 100 timesteps following a 100-timestep burn-in period,
% subset the resulting table to include only the last timestep recorded for each individual,
and calculated the variance in number of offspring across all individuals in timesteps 50-100.
All calculations were performed with information recorded in the tree sequence,
using pyslim \citep{pyslim}.

Also note that $\sigma$ controls the dispersal of offspring away from only \emph{one} parent \llabel{ll:sigma_effective}
(e.g., the seed parent for plants),
while the usual population genetics definition of dispersal distance
involves the distance to a randomly chosen one of the two parents.
(thus taking into account the distance from pollen parent as well).
This second component -- the distance between mates --
has in our simulations a distribution that is of order $\sigma$ but that depends on the population's patchiness.
If both between-mate distance and dispersal distance has variance $\sigma^2$ along each axis,
then the mean squared distance to a randomly chosen parent along that axis would be
$(\sigma^2 + 2 \sigma^2)/2 = 3 \sigma^2 /2$.
Also note that Wright's definition of neighborhood size should rightly use the \emph{effective}
dispersal distance, i.e., the mean squared displacement along an axis
between parent-child pairs found moving back along a lineage \citep{barton2002neutral}.
Nonetheless, we use $\sigma$ and $\rho$ as defined here to compute $N_W$
because these quantities are more easily observable in practice than their ``effective'' versions.
(Also, note that effective population density is lower than the actual population density,
while the $\sigma$ we use is an underestimate,
so our numbers may be close to the ``effective'' values used in theory.

@@

The effects of sampling on summary statistic estimates were summarized by testing for differences in mean (ANOVA, \citet{Rcore2018}) and variance (Levene's test, \citet{Fox2011}) across sampling strategies for each summary statistic.

@@

Run times for our model scale approximately linearly with neighborhood size (Figure \ref{fig:runtimes}), with the lowest neighborhood sizes reaching 30N generations in around a day and those with neighborhood size approaching 1,000 requiring up to three weeks of computation. As currently implemented, running simulations at neighborhood sizes more than 1,000 to coalescence is likely impractical, though running these models for more limited timescales and then ``recapitating'' the simulation using reverse-time simulation from the resulting tree sequence in msprime is possible \citep{haller2019treesequence}.  \llabel{runtimesxn}

@@

Previous studies have found that population structure and nonrandom sampling can create spurious signals of population bottlenecks when attempting to infer demographic history with microsatellite variation, summary statistics, or runs of homozygosity \citep{Chikhi2010,Stadler2009,Ptak2002,Mazet2015,leblois2006genetics}. Here we found that methods that infer detailed population trajectories through time based on the SFS and patterns of LD across the genome are also subject to this bias, with some combinations of dispersal and sampling strategy systematically inferring deep recent population bottlenecks and overestimating ancient $N_{e}$ by around a factor of 2. We were surprised to see that both stairwayplot and SMC++ can tolerate relatively strong isolation by distance -- i.e., neighborhood sizes of 20 -- and still perform well when averaging results across multiple simulations.
(However, note the high amount of between-simulation variance seen in \autoref{fig:demography_supplement}.) \llabel{ll:variance_note}
Inference in populations with neighborhood sizes over 20 was relatively unbiased unless samples were concentrated in the middle of the range (Figure \ref{fig:demography}). 
Although median demography estimates across many independent simulations were fairly accurate,
empirical work has only a single estimate to work with,
and individual model fits (Figure \ref{fig:demography_supplement})
suggest that spuriously inferred population size changes and bottlenecks are common, especially at small neighborhood sizes. 
As we will discuss below, most empirical estimates of neighborhood size, including all estimates for human populations, are large enough that population size trajectories inferred by these approaches should not be strongly affected by spatial biases created by dispersal in continuous landscapes. In contrast, \citet{Mazet2015} found that varying migration rates through time could create strong biases in inferred population trajectories from an $n$-island model with parameters relevant for human history, suggesting that changes in migration rates through time are more likely to drive variation in inferred $N_{e}$ than isolation by distance.

@@

We suggest a straightforward alternative for species in which the primary axes of population differentiation
are associated with geography (note this is likely not the case for some modern human populations): 
run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations.
If significant associations with sample locality are observed after correcting for population structure, 
the method is sensitive to false positives induced by spatial structure. This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes). This approach has recently been taken with polygenic scores for UK Biobank samples in \citet{haworth2019apparent}, finding that scores are correlated with birth location even in this relatively homogenous sample. 
Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.

@@

One facet of spatial variation that we did not address in this study is the confounding of dispersal and population density implicit in the definition of Wright's neighborhood size. Our simulations were run under constant densities, but \citet{guindon2016demographic} and \citet{Ringbauer2017} 
have shown that these parameters are identifiable under some continuous models. Similarly, though the scaling effects of dispersal we show in Figure \ref{fig:sumstats} should occur in populations of any total size, other aspects such as the number of segregating sites are also likely affected by the total landscape size (and so total census size). \llabel{landscapesizecaveat} 
Indeed, our finding that stepping-stone and continuous-space models match
in only certain aspects of genetic variation (Figure \ref{fig:grid_stats})
shows that qualitatively similar models can produce different results dependent on the specific parameterizations used.
While we believe our continuous model is a more appropriate depiction of 
many species' demographies than lattice models,
it is likely that some populations and breeding systems do more closely resemble a series of interconnected random-mating populations. As with all population models, the best approximation for any empirical system will depend on the natural history of the species in question. \llabel{bestmodelcaveat}
Much additional work remains to be done to better understand how life history, range size,
and habitat geometry interact to shape genetic variation in continuous space, which we leave to future studies.
